Realizability-Preserving DG-IMEX Method for the Two-Moment Model of Fermion Transport \tnoteref{support}

### 2018-04-19 Documents/M1/realizableFermionicM1/sections/DGIMEXscheme.tex
forgot left-hand side of equation
Cosmetic Changes
Minor adjustments to dg.tex and imex.tex.  Began work on updating DGIMEX.tex

The realizability preserving DG scheme is designed to ensure realizability of the cell averages in each element $\bK$, defined as
\begin{equation}
  \vect{\cM}_{\bK}
  =\f{1}{|\bK|}\int_{\bK}\vect{\cM}_{h}\,d\bx.  
\end{equation}
Then, with $v=1$ in Eq.~\eqref{eq:semidiscreteDG}, the stage values for the cell average in the IMEX scheme in Eq.~\eqref{eq:imexStagesRewrite} are given by
\begin{equation}
  \vect{\cM}_{\bK}^{(i)}
  =c_{i0}\,\vect{\cM}_{\bK}^{n}
  +\sum_{j=1}^{i-1}c_{ij}\,\vect{\cM}_{\bK}^{(ij)}
  +a_{ii}\,\dt\,\f{1}{\tau}\,\big(\,\vect{\eta}-\vect{\cD}\,\vect{\cM}_{\bK}^{(i)}\,\big),
  \label{eq:imexStagesCellAverage}
\end{equation}
where we have defined
\begin{equation}
  \vect{\cM}_{\bK}^{(ij)}
  =\vect{\cM}_{\bK}^{(j)}-\hat{c}_{ij}\,\dt\,\big\langle\,\nabla\cdot\vect{\cF}(\vect{\cM}_{h}^{(j)})\,\big\rangle_{\bK},
\end{equation}
and the cell average of the divergence operator is (cf. Section~\ref{sec:dg})
\begin{equation}
  \big\langle\,\nabla\cdot\vect{\cF}(\vect{\cM}_{h})\,\big\rangle_{\bK}
  =\f{1}{|\bK|}\sum_{i=1}^{d}\int_{\tilde{\bK}^{i}}
  \big(\,\widehat{\bcF}^{i}(\vect{\cM}_{h})\big|_{\xH^{i}}-\widehat{\bcF}^{i}(\vect{\cM}_{h})\big|_{\xL^{i}}\,\big)\,d\tilde{\vect{x}}^{i}.  
\end{equation}

@@

We proceed to establish conditions for which $\vect{\cM}_{\bK}^{(ij)}\in\cR$.  
To this end, to simplify notation, we consider the moments
\begin{equation}
  \vect{\cM}_{\bK}^{*}
  =\vect{\cM}_{\bK}-\hat{c}\,\dt\,\big\langle\,\nabla\cdot\vect{\cF}(\vect{\cM}_{h})\,\big\rangle_{\bK},
\end{equation}
where $\vect{\cM}_{\bK}$ is the cell average of $\vect{\cM}_{h}\in\bbV^{k}$ and $\hat{c}\ge0$.  
\begin{lemma}
  Let $\{s_{i}\}_{i=1}^{d}$ be a set of positive constants satisfying $\sum_{i=1}^{d}=1$.  
  If for each $i\in\{1,\ldots,d\}$, 
  \begin{equation}
    \vect{\Gamma}^{i}\big[\vect{\cM}_{h}\big](\tilde{\vect{x}}^{i})
    :=\f{1}{|K^{i}|}
    \Big[\,\int_{K^{i}}\vect{\cM}_{h}\,dx^{i}-\f{\hat{c}\,\dt}{s_{i}}\big(\,\widehat{\bcF}^{i}(\vect{\cM}_{h})\big|_{\xH^{i}}-\widehat{\bcF}^{i}(\vect{\cM}_{h})\big|_{\xL^{i}}\,\big)\,\Big]\in\cR,
    \label{eq:realizableGamma}
  \end{equation}
  then $\vect{\cM}_{\bK}^{*}\in\cR$.  
\end{lemma}
\begin{proof}
  It is easy to show that $\vect{\cM}_{\bK}^{*}$ can be expressed as the convex combination
  \begin{equation}
    \sum_{i=1}^{d}s_{i}\,\f{1}{|\tilde{\vect{\bK}}^{i}|}\int_{\tilde{\bK}^{i}}\vect{\Gamma}^{i}\big[\vect{\cM}_{h}\big]\,d\tilde{\vect{x}}^{i}.  
    \label{eq:cellAverageInTermsOfGamma}
  \end{equation}
  The result follows immediately.  
\end{proof}
\begin{rem}
  If a quadrature rule $\tilde{\vect{Q}}^{i}:C^{0}\to\bbR$, with positive weights, and points defined by the set $\tilde{\vect{S}}^{i}$, is used to approximate the integral over $\tilde{\bK}^{i}$ in \eqref{eq:cellAverageInTermsOfGamma}, it is sufficient for \eqref{eq:realizableGamma} to hold in the quadrature points $\tilde{\vect{S}}^{i}\subset\tilde{\bK}^{i}$.  
\end{rem}

@@

With the GL quadrature rule in \eqref{eq:quadratureRuleGL} (note $\hat{w}_{1}=\hat{w}_{N}$), we can write $\vect{\Gamma}$ as the convex combination
\begin{equation}
  \vect{\Gamma}\big[\bcM_{h}\big]
  =\sum_{q=2}^{N-1}\hat{w}_{q}\,\hat{\vect{\cM}}_{q}
  +2\,\hat{w}_{N}\,\Phi(\vect{\cM}_{\Low}^{-},\vect{\cM}_{\Low}^{+},\vect{\cM}_{\Hgh}^{-},\vect{\cM}_{\Hgh}^{+}),
  \label{eq:realizableGammaConvex}
\end{equation}
where we have defined
\begin{align}
  &\Phi(\vect{\cM}_{\Low}^{-},\vect{\cM}_{\Low}^{+},\vect{\cM}_{\Hgh}^{-},\vect{\cM}_{\Hgh}^{+}) \nonumber \\
  &\hspace{12pt}
  =\f{1}{2}\big(\,\vect{\cM}_{\Low}^{+}+\lambda\,\mathscr{F}(\vect{\cM}_{\Low}^{-},\vect{\cM}_{\Low}^{+})\,\big)
  +\f{1}{2}\big(\,\vect{\cM}_{\Hgh}^{-}-\lambda\,\mathscr{F}(\vect{\cM}_{\Hgh}^{-},\vect{\cM}_{\Hgh}^{+})\,\big),
\end{align}
with $\lambda=\hat{c}\,\dt/(s\,\hat{w}_{N}\,|K|)$.  
The following Lemma establishes sufficient conditions for realizability of $\vect{\Gamma}$, and hence $\vect{\cM}_{\bK}^{*}$.  
\begin{lemma}
  Assume that $\hat{\vect{\cM}}_{q}\in\cR$ for all $q=1,\ldots,N$ and all $\bK\in\mathscr{T}$.  
  Let the time step $\dt$ be chosen so that $\lambda\le1$.  
  Let the numerical flux be given by the Lax-Friedrichs flux in Eq.~\eqref{eq:fluxFunctionLF} with $\alpha=1$.  
  Then $\vect{\Gamma}[\bcM_{h}]\in\cR$.  
\end{lemma}
\begin{proof}
  In Eq.~\eqref{eq:realizableGammaConvex}, $\vect{\Gamma}[\vect{\cM}_{h}]$ is expressed as a convex combination.  
  By assumption, $\hat{\vect{\cM}}_{q}\in\cR$ ($q=2,\ldots,N-1$).  
  It remains to show $\Phi(\vect{\cM}_{\Low}^{-},\vect{\cM}_{\Low}^{+},\vect{\cM}_{\Hgh}^{-},\vect{\cM}_{\Hgh}^{+})\in\cR$.  
  Using the Lax-Friedrichs flux with $\alpha=1$, it is straightforward to show that
  \begin{align}
    &\Phi(\vect{\cM}_{\Low}^{-},\vect{\cM}_{\Low}^{+},\vect{\cM}_{\Hgh}^{-},\vect{\cM}_{\Hgh}^{+})
    =(1-\lambda)\,\f{1}{2}\,\big(\,\vect{\cM}_{\Low}^{+}+\vect{\cM}_{\Hgh}^{-}\,\big) \nonumber \\
    &\hspace{12pt}
    +\f{1}{2}\,\lambda\,\big(\,\Phi^{+}(\vect{\cM}_{\Low}^{-})+\Phi^{-}(\vect{\cM}_{\Hgh}^{-})\,\big)
    +\f{1}{2}\,\lambda\,\big(\,\Phi^{+}(\vect{\cM}_{\Low}^{+})+\Phi^{-}(\vect{\cM}_{\Hgh}^{+})\,\big),
    \label{eq:phiConvex}
  \end{align}
  where $\Phi^{\pm}(\vect{\cM})=\f{1}{2}\,\big(\vect{\cM}\pm\vect{e}\cdot\vect{\cF}(\vect{\cM})\big)$.  
  Since $\lambda\le1$, $\Phi$ is expressed as a convex combination of three terms.  
  By assumption, $\vect{\cM}_{\Low}^{-},\vect{\cM}_{\Low}^{+},\vect{\cM}_{\Hgh}^{-},\vect{\cM}_{\Hgh}^{+}\in\cR$, which immediately implies realizability of the first term on the right-hand side of Eq.~\eqref{eq:phiConvex}.  
  Realizability of the second and third terms follows by invoking Lemma~\ref{lem:explicitStep}.  
  This completes the proof.  
\end{proof}

>>>>

Here we proceed to develop realizability-preserving DG schemes for the two-moment model based on the IMEX schemes discussed in the previous section.  
The realizability-preserving DG scheme is designed to preserve realizability of cell averages over a time step in each element $\bK$.  
The cell average of the moments is defined as
\begin{equation}
  \vect{\cM}_{\bK}
  =\f{1}{|\bK|}\int_{\bK}\vect{\cM}_{h}\,d\bx.  
\end{equation}
Then, with $v=1$ in Eq.~\eqref{eq:semidiscreteDG}, the stage values for the cell average in the IMEX scheme in Eq.~\eqref{eq:imexStagesRewrite} are given by
\begin{equation}
  \vect{\cM}_{\bK}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\,\vect{\cM}_{\bK}^{(ij)}
  +a_{ii}\,\dt\,\f{1}{\tau}\,\big(\,\vect{\eta}-\vect{\cD}\,\vect{\cM}_{\bK}^{(i)}\,\big),
  \label{eq:imexStagesCellAverage}
\end{equation}
where we have defined
\begin{equation}
  \vect{\cM}_{\bK}^{(ij)}
  =\vect{\cM}_{\bK}^{(j)}-\hat{c}_{ij}\,\dt\,\big\langle\,\nabla\cdot\vect{\cF}(\vect{\cM}_{h}^{(j)})\,\big\rangle_{\bK},
\end{equation}
with $\cM_{\bK}^{(0)}=\cM_{\bK}^{n}$, and the cell average of the divergence operator is (cf. Section~\ref{sec:dg})
\begin{equation}
  \big\langle\,\nabla\cdot\vect{\cF}(\vect{\cM}_{h})\,\big\rangle_{\bK}
  =\f{1}{|\bK|}\sum_{i=1}^{d}\int_{\tilde{\bK}^{i}}
  \big(\,\widehat{\bcF}^{i}(\vect{\cM}_{h})\big|_{\xH^{i}}-\widehat{\bcF}^{i}(\vect{\cM}_{h})\big|_{\xL^{i}}\,\big)\,d\tilde{\vect{x}}^{i}.  
\end{equation}

@@

We proceed to establish conditions for which $\vect{\cM}_{\bK}^{(ij)}\in\cR$.  
To this end, to simplify notation, we consider the moments
\begin{equation}
  \vect{\cM}_{\bK}^{*}
  =\vect{\cM}_{\bK}-\hat{c}\,\dt\,\big\langle\,\nabla\cdot\vect{\cF}(\vect{\cM}_{h})\,\big\rangle_{\bK},
\end{equation}
where $\vect{\cM}_{\bK}$ is the cell average of $\vect{\cM}_{h}\in\bbV^{k}$ and $\hat{c}\ge0$.  
\begin{lemma}
  Let $\{s_{i}\}_{i=1}^{d}$ be a set of positive constants satisfying $\sum_{i=1}^{d}s_{i}=1$.  
  If for each $i\in\{1,\ldots,d\}$, 
  \begin{equation}
    \vect{\Gamma}^{i}\big[\vect{\cM}_{h}\big](\tilde{\vect{x}}^{i})
    :=\f{1}{|K^{i}|}
    \Big[\,\int_{K^{i}}\vect{\cM}_{h}\,dx^{i}-\f{\hat{c}\,\dt}{s_{i}}\big(\,\widehat{\bcF}^{i}(\vect{\cM}_{h})\big|_{\xH^{i}}-\widehat{\bcF}^{i}(\vect{\cM}_{h})\big|_{\xL^{i}}\,\big)\,\Big]\in\cR,
    \label{eq:realizableGamma}
  \end{equation}
  then $\vect{\cM}_{\bK}^{*}\in\cR$.  
\end{lemma}
\begin{proof}
  It is easy to show that $\vect{\cM}_{\bK}^{*}$ can be expressed as the convex combination
  \begin{equation}
    \vect{\cM}_{\bK}^{*}
    =\sum_{i=1}^{d}s_{i}\,\f{1}{|\tilde{\vect{\bK}}^{i}|}\int_{\tilde{\bK}^{i}}\vect{\Gamma}^{i}\big[\vect{\cM}_{h}\big]\,d\tilde{\vect{x}}^{i}.  
    \label{eq:cellAverageInTermsOfGamma}
  \end{equation}
  The result follows immediately.  
\end{proof}
\begin{rem}
  If a quadrature rule $\tilde{\vect{Q}}^{i}:C^{0}\to\bbR$, with positive weights, and points defined by the set $\tilde{\vect{S}}^{i}$, is used to approximate the integral over $\tilde{\bK}^{i}$ in \eqref{eq:cellAverageInTermsOfGamma}, it is sufficient for \eqref{eq:realizableGamma} to hold in the quadrature points $\tilde{\vect{S}}^{i}\subset\tilde{\bK}^{i}$.  
\end{rem}

@@

With the GL quadrature rule in \eqref{eq:quadratureRuleGL} (note $\hat{w}_{1}=\hat{w}_{N}$), we can write $\vect{\Gamma}$ as the convex combination
\begin{equation}
  \vect{\Gamma}\big[\bcM_{h}\big]
  =\sum_{q=2}^{N-1}\hat{w}_{q}\,\hat{\vect{\cM}}_{q}
  +2\,\hat{w}_{N}\,\Phi(\vect{\cM}_{\Low}^{-},\vect{\cM}_{\Low}^{+},\vect{\cM}_{\Hgh}^{-},\vect{\cM}_{\Hgh}^{+}),
  \label{eq:realizableGammaConvex}
\end{equation}
where we have defined
\begin{align}
  &\Phi(\vect{\cM}_{\Low}^{-},\vect{\cM}_{\Low}^{+},\vect{\cM}_{\Hgh}^{-},\vect{\cM}_{\Hgh}^{+}) \nonumber \\
  &\hspace{12pt}
  =\f{1}{2}\big(\,\vect{\cM}_{\Low}^{+}+\lambda\,\mathscr{F}(\vect{\cM}_{\Low}^{-},\vect{\cM}_{\Low}^{+})\,\big)
  +\f{1}{2}\big(\,\vect{\cM}_{\Hgh}^{-}-\lambda\,\mathscr{F}(\vect{\cM}_{\Hgh}^{-},\vect{\cM}_{\Hgh}^{+})\,\big),
\end{align}
with $\lambda=\hat{c}\,\dt/(s\,\hat{w}_{N}\,|K|)$.  
The following Lemma establishes sufficient conditions for realizability of $\vect{\Gamma}$, and hence $\vect{\cM}_{\bK}^{*}$.  
\begin{lemma}
  Assume that $\hat{\vect{\cM}}_{q}\in\cR$ for all $q=1,\ldots,N$ and all $\bK\in\mathscr{T}$.  
  Let the time step $\dt$ be chosen so that $\lambda\le1$.  
  Let the numerical flux be given by the Lax-Friedrichs flux in Eq.~\eqref{eq:fluxFunctionLF} with $\alpha=1$.  
  Then $\vect{\Gamma}[\bcM_{h}]\in\cR$.  
\end{lemma}
\begin{proof}
  In Eq.~\eqref{eq:realizableGammaConvex}, $\vect{\Gamma}[\vect{\cM}_{h}]$ is expressed as a convex combination.  
  By assumption, $\hat{\vect{\cM}}_{q}\in\cR$ ($q=2,\ldots,N-1$).  
  It remains to show $\Phi(\vect{\cM}_{\Low}^{-},\vect{\cM}_{\Low}^{+},\vect{\cM}_{\Hgh}^{-},\vect{\cM}_{\Hgh}^{+})\in\cR$.  
  Using the Lax-Friedrichs flux with $\alpha=1$, it is straightforward to show that
  \begin{align}
    &\Phi(\vect{\cM}_{\Low}^{-},\vect{\cM}_{\Low}^{+},\vect{\cM}_{\Hgh}^{-},\vect{\cM}_{\Hgh}^{+})
    =(1-\lambda)\,\big(\,\f{1}{2}\,\vect{\cM}_{\Low}^{+}+\f{1}{2}\,\vect{\cM}_{\Hgh}^{-}\,\big) \nonumber \\
    &\hspace{12pt}
    +\f{1}{2}\,\lambda\,\big(\,\Phi^{+}(\vect{\cM}_{\Low}^{-})+\Phi^{-}(\vect{\cM}_{\Hgh}^{-})\,\big)
    +\f{1}{2}\,\lambda\,\big(\,\Phi^{+}(\vect{\cM}_{\Low}^{+})+\Phi^{-}(\vect{\cM}_{\Hgh}^{+})\,\big),
    \label{eq:phiConvex}
  \end{align}
  where $\Phi^{\pm}(\vect{\cM})=\f{1}{2}\,\big(\vect{\cM}\pm\vect{e}\cdot\vect{\cF}(\vect{\cM})\big)$.  
  Since $\lambda\le1$, $\Phi$ is expressed as a convex combination of three terms.  
  By assumption, $\vect{\cM}_{\Low}^{-},\vect{\cM}_{\Low}^{+},\vect{\cM}_{\Hgh}^{-},\vect{\cM}_{\Hgh}^{+}\in\cR$, which immediately implies realizability of the first term on the right-hand side of Eq.~\eqref{eq:phiConvex}.  
  Realizability of the second and third terms follows by invoking Lemma~\ref{lem:explicitStep}.  
  This completes the proof.  
\end{proof}
